Learning to Synthesize Rules for Rule-based Intrusion Detectors

### 2020-03-06 paper/paper.tex
pass on discussion

\section{Network Intrusion Detection Systems (\nids)}
\label{sec:background}

@@

\sloppy \nids{} are typically categorized in two
groups~\cite{kumar2007survey}: rule-based, as briefly described on
Section~\ref{sec:intro}, and anomaly-based \nids. An
\emph{anomaly-based intrusion detector} learns usage behavior from
benign traffic and, based on that, alerts uncommon
behavior~\cite{7579764,kumar2007survey,Mitchell:2014:SID:2597757.2542049,cordy-etal-issta19}. More
precisely, an anomaly-based \nids{} flags an attack when the network
traffic manifests different characteristics compared to those of the
benign traffic. Rule-based \nids{} and anomaly-based \nids{} are
complementary. A rule-based \nids{} focuses on known attacks whereas
anomaly-based \nids{} focuses on unknown potential attacks. Rule-based
\nids\ can miss attacks as rulesets can become obsolete whereas
anomaly-based \nids\ can report false alarms as learning algorithms
that generalize observations from bening traffic are approximate.
This paper focuses on rule-based \nids\ for its tremendous popularity
in industry\Fix{cite or tone down}. Although we used Suricata, the
principles we used are general to any other signature-based
\nids~(\eg{}, Snort~\cite{snort}).

\subsection{Suricata Rules}
\label{sec:example-suricata-rules}

Figure~\ref{fig:pingscan-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{} maintained by
the Open Information Security Foundation (OISF)~\cite{oisf}.  A
Suricata rule is divided in three parts---action, header, and rule
options~\cite{suri-rule-format}. The \emph{action} appears as the
first word in the rule description (an \CodeIn{alert}, in this
case). An action denotes the task that needs to be executed if the
rule pattern is satisfied.  In this example, an alert message will be
sent to system administrators if the rule pattern is
satisfied\footnote{Technically, an IPS (Intrusion Prevention System)
  differs from an IDS in the action taken when an intrusion is
  observed. IPS systems block malicous traffic whereas IDS system only
  alert them. However, IDS and IPS do not differ in the approaches
  used to detect the intrusion~\cite{ids-ips}.}. The
\emph{header} comes after the action in the rule description. It
restricts the information flow covered by the rule. The header of this
rule is \CodeIn{icmp any any -> any any}. It instructs Suricata to
inspect \CodeIn{icmp} traffic flowing from any machine and port to any
machine and port.\MyComment{ The variables \CodeIn{\$HOME\_NET} and
  \CodeIn{\$EXTERNAL\_NET} are configurable.} Finally, \emph{rule
  options} come after the header. It is a semi-colon-separated
sequence of key-value pairs. Rule options serve to document the
analyzed traffic (\eg{}, ``msg'', ``classtype'', and ``sid'') and to
characterize the attack pattern (\eg, ``flags'', ``threshold'', and
``content''). \tname{} produces options of the second kind, which are
\emph{relevant} to capture the attack. For example, the option
\CodeIn{content:<arg>} checks if the argument is present in the
payload of a message.

\vspace{-1ex}
\begin{table}[h!]
  \footnotesize
  \caption{\label{table:rules}Example of non-documentation options of Suricata.}
  \vspace{-2ex}
  \centering
  \begin{tabular}{clp{6cm}}
    \toprule
    \multicolumn{1}{c}{\#} & \multicolumn{1}{c}{Name} &  \multicolumn{1}{c}{Description}\\
    \midrule     
    1 & dsize & matches on the size of the packet payload\\
    2 & itype &  matches on a specific ICMP type\\
    3 & icode & matches on a specific ICMP code\\
    4 & icmp\_seq  & checks a ICMP sequence number\\
    5 & icmp\_id & matches on specific ICMP id-values\\
    6 & window & checks the size of the TCP window\\
    7 & flags & matches on TCP flags\\
    8 & fragbits & checks if the frag. or reserved bits are set in the IP header\\
    9 & threshold & minimum threshold for a rule before it generates alerts\\
    10 & content & checks if argument is present on the payload\\
    \bottomrule
  \end{tabular}
\end{table}
%\vspace{-1ex}

@@

\newcommand{\textRQone}{How well \tname\ ranks the golden rule?}
\vspace{0.2cm}
\begin{itemize}[leftmargin=*,label={}]
\item{\textbf{RQ1.}} \textRQone
\end{itemize}

@@

\newcommand{\textRQtwo}{How important are the heuristic functions
  to \tname's performance?}
\vspace{0.2cm}
\begin{itemize}[leftmargin=*,label={}]
\item{\textbf{RQ2.}} \textRQtwo\
\end{itemize}

@@

\newcommand{\textRQthree}{How sensitive \tname\ is to the amount of
  positive examples?}
\vspace{0.2cm}
\begin{itemize}[leftmargin=*,label={}]
\item{\textbf{RQ3.}} \textRQthree\
\end{itemize}

@@

\begin{center}
\begin{tcolorbox}[enhanced,width=3.3in,center upper,drop shadow southwest,sharp corners]
\tname\ successfully synthesizes the golden rules in all cases and
that rule is ranked among the first top five rules in
\percTopFiveRanking\ of the cases.
\end{tcolorbox}
\end{center}

@@

\begin{figure}[h!]
  \centering
  \begin{subfigure}{.2\textwidth}
    \centering
    \includegraphics[scale=0.4,trim=20 0 130 0,clip]{R/idqaccess/idqaccess.pdf}
    \caption{idqaccess}
  \end{subfigure}%
  \begin{subfigure}{.2\textwidth}
    \centering
    \includegraphics[scale=0.4,trim=15 0 130 0,clip]{R/scripttag/scripttag.pdf}
    \caption{scripttag}    
  \end{subfigure}%
\vspace{-2ex}  
\caption{Impact of number of positive examples.}
\label{fig:impact-number-attacks}
\vspace{-2ex}
\end{figure}

@@

\section{Discussion}

\subsection{Sample of other attacks}

\subsubsection{Remote Code Execution (RCE)}
\label{sec:rce}
\label{sec:jboss}
\label{sec:content-example}

%% Adaptacao do texto copiado de:  https://www.solarwindsmsp.com/blog/remote-code-execution

RCE refers to the family of attacks where hackers exploit a network
vulnerability to run arbitrary code---that could damage the system or
steal sensitive information---on a remote machine. JBoss~\cite{jboss}
is popular open source Java application server. This attacks enables
remote execution of code on the JBoss application server.
\Mar{@Lucas, por favor, observe o padrao usado nos exemplos
  anteriores. eh preciso explicar em alto nivel o ataque antes de
  explicar a regra!}

@@

\begin{figure}[H]
  \lstinputlisting[language=C,numbers=none,keywords={content,flow}]{adaptor-golden-rule.suricata}
  \caption{\label{fig:rule-jboss}Suricata rule for JBoss Remote Code Execution.}
\end{figure}

@@

Figure~\ref{fig:rule-jboss} shows a Suricata rule that captures a
JBoss RCE attack.  The option \CodeIn{flow: established, to\_server;}
indicates that a rule match will only occur on packets that (1) flow
on an already-established connection and that (2) flow to the server
as opposed to the client. For the content options the modifiers
\CodeIn{nocase} and \CodeIn{http\_uri} indicate, respectively, that
the matching not be case sensitive and that the string should be
located at the packet field ``Request URI''. \Mar{Lucas, ``nocase''
  tambem eh considerado modificador? Isto nao eh confuso? Na secao
  4.3.4 consideramos modificadores como sinonimo para qualificador de
  campo, nao?}\Luc{Sim, ``nocase'' define que a busca pela string nao deve ser case sensitive. Em 4.3.4 estamos considerando apenas os modificadores que definem posições/campos de pacotes http.}

\tname{} reproduces the malicious JBoss attack using the traffic from
a Nikto script~\cite{nikto}. However, similar JBoss attack is also
available from the Metasploit database~\cite{metasploit}. As in the
other examples, \tname{} uses, as additional inputs, benign traffic
from \Fix{...} and a set of existing Suricata rules from
\Fix{...}. \tname{} initially produces one seed rule containing
\Fix{xx} options, by analyzing the malicious traffic alone. Then, it
uses that seed rule to derive \Fix{yy} rules, each rule containing
subsets of the option in the initial rule. Finally, it ranks that
list. The list below shows the top-5 rules that \tname{}
produces.\Mar{<- check/revise}

\Fix{list rules}

Note that the golden rule is in position \Fix{...} on this
case. \Mar{explain syntactical difference, if exist}

\subsubsection{Denial-of-Service}
\label{sec:dos}

\Mar{move from above --> We used the following hping3 command to
  reproduce that attack: \CodeIn{hping3 -d 80 -w 512 -S -p 80 --flood
    --rand-source <IP>}.}

The SYN Flood attack is a denial-of-service attack that exploits a
vulnerability in the tcp/ip handshake
to establish a tcp connection~\cite{cloudfare-synflood}. The handshake
works as follows in normal circumstances. First, a client sends a SYN
packet to the server, requesting a connection. Second, the server
responds with a SYN-ACK packet to the client. Third, the client
responds with an ACK message and the connection is established. Aware
of the protocol, an attacker sends multiple SYN packets to different
ports of a server, often using fake IP addresses. Then, after the
server responds with a SYN-ACK packet, the client keeps sending other
SYN packets to avoid the connection to time out. Without proper
protection, the server accepts these malicious requests and eventually
legitimate requests cannot be satisfied due to resource exhaustion.

@@

Figure~\ref{fig:synflood-example} shows a Suricata rule that captures
this attack. This rule was obtained from the Emerging Threats Open
Ruleset~\cite{emerging-threats-open}. The relevant options in the
rule appear in bold---the option \CodeIn{flags: S,12}, identifies a
SYN packet in a tcp packet and the option \CodeIn{threshold: type
  both, track by\_dst, count 5000, seconds 5} indicates that a high
volume of such packets should be requested in a short period of
time. The other options are for documentation purposes.

...........
In this case, however, that information alone is
insufficient as the malicious traffic is characterized as a sequence
of messages. 
...........

%% \Mar{...and where did they come from?} \Luc{Sim, floods podem ser
%% compostos por milhares de pacotes, um dos data sets q encontrei
%% possui mais de 500k pacotes de flood. Os data set  s foram encontrados na internet.}

\begin{figure}[h]
  \vspace{-2ex}
  \lstinputlisting[language=C,numbers=none,frame=none,keywords={flags,threshold,window}]{synflood.suricata.synth}
%  \caption{Suricata rule for SYN Flood Attacks.}
  %  \label{fig:synflood-example-synt}
  \vspace{-2ex}  
\end{figure}

%% \Mar{@Lucas/Guilheme, há algo o que remover? talvez ``window:64''?
%%   Qual a correspondencia de ``flags:S,12'' com ``flags:S''? Por favor,
%%   respondam isto logo.}  \Gui{O window:64 seria removido
%%   sim.}\Mar{ok. vou explicar isto.}
  %% \Luc{O segundo campo da opção flags define flags que serão
  %%   ignoradas. Nesse caso a regra procura por pacotes onde apenas a
  %%   flag S está 'setada', independentemente do valor das flags 1 e
  %%   2. Portanto a regra gerada pode capturar coisas diferentes da
  %%   regra original. A inferencia do "12" n pode ser feita apenas a
  %%   partir do pacote de entrada, precisariamos tratar
  %%   isso.}\Mar{vc. pode explicar isto **mais detalhadamente** \Luc{a opção flags pode ter dois campos separados por virgula (ex: flags:S,12). o primeiro campo indica quais flags devem estar ativas em um pacote para que o suricata gere um alerta. se uma ou mais flags forem irrelevantes para a detecção de um determinado ataque (ou seja, se n importa se a flag está ou não ativa), elas serão adicionadas ao segundo campo.} E (1)
  %%   explicar que isto eh uma limitacao \Luc{n entendi exatamente oq explicar aqui.} e (2) indicar a possivel consequencia
  %%   disto? \Luc{a consequencia sao possiveis falsos negativos. ao não incluir flags no segundo campo podemos descartar pacotes por conta da presença de flags q n são relevantes (ou seja, vamos levar em conta flags q n deveriam interferir na detecção).}}

\noindent

.....

 The list below shows the rule options for the top-5 rules
\tname{} produces for this attack.\Mar{algum problema em fazer isto?
  exemplos concretos ajudam muito!} \Luc{No caso desse ataque, a regra inicial possui apenas duas opcoes (sem contar o threshold). Entao existem apenas 3 combinacoes possiveis, dessas 3 apenas 2 sao capazes de isolar o ataque do trafego normal. Portanto, o top-2 eh:
  \begin{figure}[h]
    \vspace{-2ex}
    \lstinputlisting[language=C,numbers=none,frame=none,keywords={window,flags, threshold}]{top-synflood.rules}
    \vspace{-3ex}  
  \end{figure}

   }\Mar{Lucas, eu nao entendo porque vc. desconsiderou threshold da
  discussao (afinal, a gente destaca thrshold como relevante no inicio
  desta secao). tambem nao entendo porque window---que a gente disse
  acima que *nao* era relevante---aparece nas duas regras
  geradas. enfim, isto esta inconsistente com nosso discurso.}
  \Luc{Coloquei o threshold em destaque tambem. O window nas duas regras foi um erro, corrigi.}

\vspace{1ex}
\noindent
\textbf{Differences.}~Note that there are differences between the
options produced in the first step of \tname{} and the options from
the golden rule, as per Figure~\ref{fig:synflood-example}. The option
\CodeIn{flags} does not mention parameters \CodeIn{1 2}, there is an
extra option \CodeIn{window}, and there are differences in the option
\CodeIn{threshold}. Considering the option \CodeIn{flags}, the first
parameter of that option indicates what flags in the packet must be
set while the second parameter indicates non-mandatory flags (\ie{},
``don't cares''). More specifically, the option \CodeIn{flags: S}
indicates that the flag ``S'' must be set and all flags not mentioned
(including ``1'' and ``2'') must be unset for the \nids\ to capture
the traffic. Consequently, that choice of option alone could result in
false negatives and false positives. An attack that satisfies all
other conditions in a rule and has flags ``1'' or ``2'' set would be
captured by the original rule but not captured by the inferred rule---a case of
false negative. Likewise, an attack that satisfies all conditions and
have flags ``1'' and ``2'' unset would be captured by the inferred
rule but not captured by the original rule---a case of false positive. The set of
positive examples is to blame in this case. To mitigate this issue,
the input messages should have covered cases with options ``1'' or
``2'' set. Considering option \CodeIn{window: 512}, \tname{} is able
to generate rules without that option, as discussed above. Considering
option \CodeIn{threshold}, the values used in the golden rule are
domain-specific. \tname{} uses default values for these options. We
conjecture that the system administrator would be able to configure
this option.

%% \begin{table}[h]
%%   \caption{\label{table:rules}Rule evolution}  
%%   \centering
%%   \begin{tabular}{lllll}
%%     \toprule
%%     \multicolumn{1}{c}{Iteration} & \multicolumn{1}{c}{Rule} \\
%%     \midrule     
%%     0 & (ack:0; seq:0; window:0; flags:F;)\\
%%     5 & (window:0; flags:F;)\\
%%     68 & (window:64; flags:F;)\\
%%     73 & (window:64; flags:S,12;)\\
%%     94 & (window:64; flags:S,12; threshold: type both, track by\_dst, count 20, seconds 1;)\\
%%     \bottomrule
%%   \end{tabular}
%% \end{table}

>>>>

\section{\nids}
\label{sec:background}

@@

\sloppy Network Intrusion Detection Systems (\nids{}) are typically categorized in two
groups~\cite{kumar2007survey}: rule-based, as briefly described on
Section~\ref{sec:intro}, and anomaly-based \nids. An
\emph{anomaly-based intrusion detector} learns usage behavior from
benign traffic and, based on that, alerts uncommon
behavior~\cite{7579764,kumar2007survey,Mitchell:2014:SID:2597757.2542049,cordy-etal-issta19}. More
precisely, an anomaly-based \nids{} flags an attack when the network
traffic manifests different characteristics compared to those of the
benign traffic. Rule-based \nids{} and anomaly-based \nids{} are
complementary. A rule-based \nids{} focuses on known attacks whereas
anomaly-based \nids{} focuses on unknown potential attacks. Rule-based
\nids\ can miss attacks as rulesets can become obsolete whereas
anomaly-based \nids\ can report false alarms as learning algorithms
that generalize observations from bening traffic are approximate.
This paper focuses on rule-based \nids\ for its tremendous popularity
in industry\Fix{cite or tone down}. Although we used Suricata, the
principles we used are general to any other signature-based
\nids~(\eg{}, Snort~\cite{snort}).

%\vspace{-1ex}
\subsection{Suricata Rules}
\label{sec:example-suricata-rules}

Figure~\ref{fig:pingscan-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{} maintained by
the Open Information Security Foundation (OISF)~\cite{oisf}.  A
Suricata rule is divided in three parts---action, header, and rule
options~\cite{suri-rule-format}. The \emph{action} appears as the
first word in the rule description (an \CodeIn{alert}, in this
case). An action denotes the task that needs to be executed if the
rule pattern is satisfied.  In this example, an alert message will be
sent to system administrators if the rule pattern is
satisfied\footnote{IPS (Intrusion Prevention System) and IDS differ in
  the action taken when an intrusion is observed. IPS systems block
  malicous traffic whereas IDS systems only alert them. However, IDS
  and IPS are similar in the approaches used to detect
  intrusions~\cite{ids-ips}.}. The \emph{header} comes after the
action in the rule description. It restricts the information flow
covered by the rule. The header of this rule is \CodeIn{icmp any any
  -> any any}. It instructs Suricata to inspect \CodeIn{icmp} traffic
flowing from any machine and port to any machine and port.\MyComment{
  The variables \CodeIn{\$HOME\_NET} and \CodeIn{\$EXTERNAL\_NET} are
  configurable.} Finally, \emph{rule options} come after the
header. It is a semi-colon-separated sequence of key-value pairs. Rule
options serve to document the analyzed traffic (\eg{}, ``msg'',
``classtype'', and ``sid'') and to characterize the attack pattern
(\eg, ``flags'', ``threshold'', and ``content''). \tname{} produces
options of the second kind, which are \emph{relevant} to capture the
attack. For example, the option \CodeIn{content:<arg>} checks if the
argument is present in the payload of a message.

\vspace{-1ex}
\begin{table}[t!]
  \footnotesize
  \caption{\label{table:rules}Example of non-documentation options of Suricata.}
  \vspace{-2ex}
  \centering
  \begin{tabular}{clp{6cm}}
    \toprule
    \multicolumn{1}{c}{\#} & \multicolumn{1}{c}{Name} &  \multicolumn{1}{c}{Description}\\
    \midrule     
    1 & dsize & matches on the size of the packet payload\\
    2 & itype &  matches on a specific ICMP type\\
    3 & icode & matches on a specific ICMP code\\
    4 & icmp\_seq  & checks a ICMP sequence number\\
    5 & icmp\_id & matches on specific ICMP id-values\\
    6 & window & checks the size of the TCP window\\
    7 & flags & matches on TCP flags\\
    8 & fragbits & checks if the frag. or reserved bits are set in the IP header\\
    9 & threshold & minimum threshold for a rule before it generates alerts\\
    10 & content & checks if argument is present on the payload\\
    \bottomrule
  \end{tabular}
  \vspace{-3ex}  
\end{table}

@@

\newcommand{\textRQone}{How well \tname\ ranks the golden rule?}
%\vspace{0.2cm}
\begin{itemize}[leftmargin=*,label={}]
\item{\textbf{RQ1.}} \textRQone
\end{itemize}

@@

\newcommand{\textRQtwo}{How important are the heuristic functions
  to \tname's performance?}
%\vspace{0.2cm}
\begin{itemize}[leftmargin=*,label={}]
\item{\textbf{RQ2.}} \textRQtwo\
\end{itemize}

@@

\newcommand{\textRQthree}{How sensitive \tname\ is to the amount of
  positive examples?}
%\vspace{0.2cm}
\begin{itemize}[leftmargin=*,label={}]
\item{\textbf{RQ3.}} \textRQthree\
\end{itemize}

@@

\begin{center}
\begin{tcolorbox}[enhanced,width=3.3in,center upper,drop shadow southwest,sharp corners]
\tname\ successfully synthesizes the golden rules in all cases and
that rule is ranked among the top five rules in
\percTopFiveRanking\ of the cases.
\end{tcolorbox}
\end{center}

@@

\begin{figure}[t!]
  \centering
  \begin{subfigure}{.2\textwidth}
    \centering
    \includegraphics[scale=0.4,trim=20 0 130 0,clip]{R/idqaccess/idqaccess.pdf}
    \caption{idqaccess}
  \end{subfigure}%
  \begin{subfigure}{.2\textwidth}
    \centering
    \includegraphics[scale=0.4,trim=15 0 130 0,clip]{R/scripttag/scripttag.pdf}
    \caption{scripttag}    
  \end{subfigure}%
\vspace{-2.5ex}  
\caption{Impact of number of positive examples.}
\label{fig:impact-number-attacks}
\vspace{1ex}
\end{figure}

@@

\section{Discussion of other attacks}
\label{sec:rce}
\label{sec:jboss}
\label{sec:content-example}
\label{sec:dos}

This section elaborates how \tname\ performs on two other attacks.

\textbf{Remote Code Execution (RCE).}  RCE refers to the family of
attacks where hackers exploit a network vulnerability to run arbitrary
code on a remote machine. The code to be executed could damage the
system or steal sensitive information. JBoss~\cite{jboss} is a popular
open source Java application server. This attack enables remote
execution of code on the JBoss application server.

%\Mar{@Lucas, por
%  favor, observe o padrao usado nos exemplos anteriores. eh preciso
%  explicar em alto nivel o ataque antes de explicar a regra!}

@@

\begin{figure}[H]
  \lstinputlisting[language=C,numbers=none,keywords={content,flow}]{adaptor-golden-rule.suricata}
  \vspace{-2ex}  
  \caption{\label{fig:rule-jboss}Suricata rule for JBoss Remote Code
    Execution.}
  \vspace{-3ex}
\end{figure}

@@

Figure~\ref{fig:rule-jboss} shows a Suricata rule that captures a
JBoss RCE attack.  The option \CodeIn{flow: established, to\_server;}
indicates that a rule match will only occur on packets that (1) flow
on an already-established connection and that (2) flow to the server
as opposed to the client. For the content options the modifiers
\CodeIn{nocase} and \CodeIn{http\_uri} indicate, respectively, that
the matching is not case sensitive and that the string should be
located at the packet field ``Request URI''. We reproduced the
malicious JBoss attack using the traffic from a Nikto
script~\cite{nikto}. Similar attacks are also available on the
Metasploit database~\cite{metasploit}. This is a case where
\tname\ performed poorly. Of the 228 plausible rules it generated for
this attack, the ground truth was located at position 45 in the ranked
list. The reason for that is that \Fix{...xxx...}

%% \Mar{move from above --> We used the following hping3 command to
%%   reproduce that attack: \CodeIn{hping3 -d 80 -w 512 -S -p 80 --flood
%%     --rand-source <IP>}.}

\textbf{Denial-of-Service.}~
The SYN Flood attack is a denial-of-service attack that exploits a
vulnerability in the tcp/ip handshake
to establish a tcp connection~\cite{cloudfare-synflood}. The handshake
works as follows in normal circumstances. First, a client sends a SYN
packet to the server, requesting a connection. Second, the server
responds with a SYN-ACK packet to the client. Third, the client
responds with an ACK message and the connection is established. Aware
of the protocol, an attacker sends multiple SYN packets to different
ports of a server, often using fake IP addresses. Then, after the
server responds with a SYN-ACK packet, the client keeps sending other
SYN packets to avoid the connection to time out. Without proper
protection, the server accepts these malicious requests and eventually
legitimate requests cannot be satisfied due to resource exhaustion.

@@

Figure~\ref{fig:synflood-example} shows a Suricata rule that captures
this attack. This rule was obtained from the Emerging Threats Open
Ruleset~\cite{emerging-threats-open}. The relevant options in the
rule appear in bold---the option \CodeIn{flags: S,12}, identifies a
SYN packet in a tcp packet and the option \CodeIn{threshold: type
  both, track by\_dst, count 5000, seconds 5} indicates that a high
volume of such packets should be requested in a short period of
time. The other options are for documentation purposes. This is a case
where \tname\ works very well, reporting the golden rule at the top
of the ranking.
